Athanasian Creed
Whoever desires to be saved must, above all, hold the catholic faith. Whoever does not 
keep it whole and undefiled will without doubt perish eternally. 
 
And the catholic faith is this, that we worship one God in Trinity and Trinity in Unity, neither 
confusing the persons nor dividing the substance. For the Father is one person, the Son is 
another, and the Holy Spirit is another. But the Godhead of the Father and of the Son and of 
the Holy Spirit is one: the glory equal, the majesty coeternal. Such as the Father is, such is 
the Son, and such is the Holy Spirit: the Father uncreated, the Son uncreated, the Holy Spirit 
uncreated; the Father infinite, the Son infinite, the Holy Spirit infinite; the Father eternal, 
the Son eternal, the Holy Spirit eternal. And yet there are not three Eternals, but one 
Eternal, just as there are not three Uncreated or three Infinites, but one Uncreated and one 
Infinite. In the same way, the Father is almighty, the Son almighty, the Holy Spirit almighty; 
and yet there are not three Almighties, but one Almighty. So the Father is God, the Son is 
God, the Holy Spirit is God; and yet there are not three Gods, but one God. So the Father is 
Lord, the Son is Lord, the Holy Spirit is Lord; and yet there are not three Lords, but one Lord. 
Just as we are compelled by the Christian truth to acknowledge each distinct person as God 
and Lord, so also are we prohibited by the catholic religion to say that there are three Gods 
or Lords. 
 
The Father is not made nor created nor begotten by anyone. The Son is neither made nor 
created, but begotten of the Father alone. The Holy Spirit is of the Father and of the Son, 
neither made nor created nor begotten, but proceeding. Thus, there is one Father, not three 
Fathers; one Son, not three Sons; one Holy Spirit, not three Holy Spirits. And in this Trinity 
none is before or after another; none is greater or less than another; but the whole three 
persons are coeternal with each other and coequal, so that in all things, as has been stated 
above, the Trinity in Unity and Unity in Trinity is to be worshiped. Therefore, whoever 
desires to be saved must think thus about the Trinity. 
 
But it is also necessary for everlasting salvation that one faithfully believe the incarnation of 
our Lord Jesus Christ. Therefore, it is the right faith that we believe and confess that our 
Lord Jesus Christ, the Son of God, is at the same time both God and man. He is God, 
begotten from the substance of the Father before all ages; and He is man, born from the 
substance of His mother in this age: perfect God and perfect man, composed of a rational 
soul and human flesh; equal to the Father with respect to His divinity, less than the Father 
with respect to His humanity. Although He is God and man, He is not two, but one Christ: 
one, however, not by the conversion of the divinity into flesh, but by the assumption of the 
humanity into God; one altogether, not by confusion of substance, but by unity of person. 
For as the rational soul and flesh is one man, so God and man is one Christ, who suffered for 
our salvation, descended into hell, rose again the third day from the dead, ascended into 
heaven, and is seated at the right hand of the Father, God Almighty, from whence He will 
come to judge the living and the dead. At His coming all people will rise again with their 
bodies and give an account concerning their own deeds. And those who have done good will 
enter into eternal life, and those who have done evil into eternal fire. 
 
This is the catholic faith; whoever does not believe it faithfully and firmly cannot be saved.
